The \dealii Library, Version 9.2

### 2020-05-10 9.2/paper.tex
Add sections: ca, pft, mf

@@

\hypersetup{
  pdfauthor={
     Daniel Arndt,
     Wolfgang Bangerth,
%     Thomas C. Clevenger,
%     Denis Davydov,
%     Marc Fehling,
%     Daniel Garcia-Sanchez,
%     Graham Harper,
    Timo Heister,
%     Luca Heltai,
    Martin Kronbichler,
%     Ross Maguire Kynch,
%     Matthias Maier,
%     Jean-Paul Pelteret,
%     Bruno Turcksin,
%     David Wells
  },
  pdftitle={The deal.II Library, Version 9.2, 2020},
}

@@

% %
% \author[8]{Luca~Heltai}
% \affil[8]{SISSA,
%   International School for Advanced Studies,
%   Via Bonomea 265,
%   34136, Trieste, Italy.
%   {\texttt{luca.heltai@sissa.it}}}
%
 \author[9]{Martin~Kronbichler}
 \affil[9]{Institute for Computational Mechanics,
   Technical University of Munich,
   Boltzmannstr.~15, 85748 Garching, Germany.
   {\texttt{kronbichler@lnm.mw.tum.de}}}
%
% \author[10]{Ross~Maguire~Kynch}
% \affil[10]{Zienkiewicz Centre for Computational Engineering,
%   College of Engineering, Swansea University,
%   Bay Campus, Fabian Way, Swansea SA1 8EN, Wales, UK.
%   {\texttt{rkynch@gmail.com}}}
%
\author[11]{Matthias~Maier}
\affil[11]{Department of Mathematics,
  Texas A\&M University,
  3368 TAMU,
  College Station, TX 77845, USA.
  {\texttt{maier@math.tamu.edu}}}
%
% \author[4]{Jean-Paul~Pelteret}
%
% \author[*1]{Bruno~Turcksin}
%
% \author[12]{David Wells}
% \affil[12]{Department of Mathematics, University of North Carolina,
%   Chapel Hill, NC 27516, USA.
%   {\texttt{drwells@email.unc.edu}}}

@@

@@

In addition, there are seven new tutorial programs:
\begin{itemize}
\item \texttt{step-47}
\item \texttt{step-50}
\item \texttt{step-58}
\item \texttt{step-65}
\item \texttt{step-67}
\item \texttt{step-69}
\item \texttt{step-70}
\end{itemize}

@@

The 9.2 release includes
\href{https://dealii.org/developer/doxygen/deal.II/changes_between_9_1_1_and_9_2_0.html}
     {around 15 incompatible changes}; see \cite{changes92}. The majority of these changes
should not be visible to typical user codes; some remove previously
deprecated classes and functions; and the majority change internal
interfaces that are not usually used in external
applications. However, some are worth mentioning:
\begin{itemize}
\item
% \item The \texttt{VectorView} class was removed. We recommend either copying the
%       vector subset into a \texttt{Vector} or using a \texttt{BlockVector}.
% \item The function \texttt{Subscriptor::subscribe()}, used through the
%   \texttt{SmartPointer} class, now requires a pointer to a
%       \texttt{std::atomic<bool>} that tracks whether or not the pointer to the
%       subscribed-to object is still valid.
% \item The \texttt{ConstraintMatrix} class gained a template parameter for the scalar
%       type and was been renamed \texttt{AffineConstraints}. Several methods that
%       take vectors or matrices as arguments,
%       such as \texttt{AffineConstraints::distribute\_local\_to\_global()},
%       now require that all matrix and vector arguments have matching number
%       types.
% \item Similarly, the functions \texttt{create\_mass\_matrix} and
%       \texttt{create\_boundary\_mass\_matrix} in the \texttt{MatrixCreator}
%       namespace no longer
%       support matrix and vector objects of different types.
\end{itemize}

>>>>

%\renewcommand{\baselinestretch}{2.0}

@@

\hypersetup{
  pdfauthor={
     Daniel Arndt,
     Wolfgang Bangerth,
%     Thomas C. Clevenger,
%     Denis Davydov,
%     Marc Fehling,
%     Daniel Garcia-Sanchez,
%     Graham Harper,
    Timo Heister,
%     Luca Heltai,
    Martin Kronbichler,
    Peter Munch,
%     Ross Maguire Kynch,
%     Matthias Maier,
%     Jean-Paul Pelteret,
%     Bruno Turcksin,
%     David Wells
  },
  pdftitle={The deal.II Library, Version 9.2, 2020},
}

@@

% %
% \author[8]{Luca~Heltai}
% \affil[8]{SISSA,
%   International School for Advanced Studies,
%   Via Bonomea 265,
%   34136, Trieste, Italy.
%   {\texttt{luca.heltai@sissa.it}}}
%
 \author[9]{Martin~Kronbichler}
 \affil[9]{Institute for Computational Mechanics,
   Technical University of Munich,
   Boltzmannstr.~15, 85748 Garching, Germany.
   {\texttt{kronbichler/munch@lnm.mw.tum.de}}}
%
% \author[10]{Ross~Maguire~Kynch}
% \affil[10]{Zienkiewicz Centre for Computational Engineering,
%   College of Engineering, Swansea University,
%   Bay Campus, Fabian Way, Swansea SA1 8EN, Wales, UK.
%   {\texttt{rkynch@gmail.com}}}
%
\author[11]{Matthias~Maier}
\affil[11]{Department of Mathematics,
  Texas A\&M University,
  3368 TAMU,
  College Station, TX 77845, USA.
  {\texttt{maier@math.tamu.edu}}}

  \author[9,13]{Peter Munch}

 \affil[13]{Institute of Materials Research, Materials Mechanics, 
 Helmholtz-Zentrum Geesthacht, 
 Max-Planck-Str. 1, 21502 Geesthacht, Germany.
   {\texttt{peter.muench@hzg.de}}}

%
% \author[4]{Jean-Paul~Pelteret}
%
% \author[*1]{Bruno~Turcksin}
%
% \author[12]{David Wells}
% \affil[12]{Department of Mathematics, University of North Carolina,
%   Chapel Hill, NC 27516, USA.
%   {\texttt{drwells@email.unc.edu}}}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Improved large-scale performance}
\label{subsec:performance}

Large-scale simulations with 304,128 cores have revealed bottlenecks in release 
9.1 during initialization due to the usage of expensive collective operations 
like \texttt{MPI\_Allgather()} and \texttt{MPI\_\allowbreak Alltoall()}. These 
operations are used to retrieve or even store information about all processes. 
For example, all processes have stored in  an array the number of degrees of 
freedom each process owns. This information is in particular needed to set up 
the  \texttt{Utilities::MPI::Partitioner} class, which contains the 
point-to-point communication pattern for vector ghost-value updates and 
compressions. In release 9.2, we have removed such arrays and have replaced 
the \texttt{MPI\_Allgather}/\allowbreak\texttt{MPI\_\allowbreak Alltoall} 
function calls by consensus algorithms~\cite{hoefler2010scalable}, which can be 
found in the namespace \texttt{Utilities::\allowbreak MPI::\allowbreak ConsensusAlgorithms}: now, only the locally relevant information is computed 
(and recomputed) when needed, using these algorithms.

Consensus algorithms are algorithms dedicated to efficient dynamic-sparse 
communication patterns. In this context, the term ``dynamic-sparse'' means 
that by the time this function is called, the other processes do not know 
yet that they have to answer requests and
each process only has to communicate with a small subset of processes of the 
MPI communicator. We provide two flavors of the consensus algorithm: the two-step 
approach \texttt{ConsensusAlgorithms::PEX} and the \texttt{ConsensusAlgorithms::NBX}, 
which uses only point-to-point communications and a single \texttt{MPI\_IBarrier()}. 
The class \texttt{ConsensusAlgorithms::Selector} selects one of the two previous 
algorithms, depending on the number of processes.

Due to the excellent scalability of the consensus algorithms, users are encouraged 
to use them for their own dynamic-sparse problems by providing a list of target 
processes and pack/unpack routines either by implementing the interface 
\texttt{ConsensusAlgorithms::Process} or by providing \texttt{std::function} 
objects to \texttt{ConsensusAlgorithms::AnonymousProcess}.

The \texttt{ConsensusAlgorithms} are used by now internally in many places. These 
places are appropriate starting points for users for their own application of the 
\texttt{ConsensusAlgorithms} infrastructure.
For example, to set up the partitioners, the new function 
\texttt{compute\_index\_owner() } is used: given an index set containing the 
locally owned indices and an index set containing the ghost indices, it returns 
the owner of the ghost indices. Consensus algorithms are used now also in the 
functions \texttt{compute\_point\_to\_point\_communication\_pattern()} and 
\texttt{compute\_\allowbreak n\_\allowbreak point\_\allowbreak to\_\allowbreak point\_\allowbreak communications()}. 
Furthermore, it is utilized in the class \texttt{NoncontiguousPartitioner} to 
efficiently permute distributed solution vectors globally in an arbitrary order, e.g., 
to interface with external libraries that prescribe a certain partitioning and 
padding of the data.

By replacing the collective communications during setup and removing the arrays 
that contain information for each process (enabled by the application of consensus 
algorithms and other modifications---a full list of modifications leading to this 
improvement can be found online), we were able to significantly improve the setup 
time for large-scale simulations and to solve a Poisson problem with multigrid 
with 12T unknowns. 
Figure~\ref{} compares the timings of a simulation (incl. setup) with the 
previous release 9.1 and with the current release 9.2. 
{\color{red}TODO[Peter/Martin] description of the results}

The new code has been also applied to solve problems with adaptively refined 
meshes with more than 4B unknowns. {\color{red}TODO[Timo]}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{A new fully distributed triangulation class}
\label{subsec:pft}

By release 9.1, \texttt{deal.II} had three types of triangulation classes: the 
serial triangulation class \texttt{Triangulation} as well as the parallel 
triangulation classes \texttt{parallel::shared::Triangulation} and \texttt{parallel::distributed::Triangulation}. The latter builds around a serial 
triangulation and uses \texttt{p4est} as an oracle during adaptive mesh refinement. 
All these triangulation classes have in common that the coarse grid is shared by 
all processes and the actual mesh used for computations is constructed by repeated 
local and/or global refinement, which adapts nicely to curved boundaries described 
by the \texttt{Manifold} class. However, this way to construct a computational 
mesh has its limitations in industrial applications where, often, the mesh comes 
from an external CAD program in the form of a file that already contains millions 
or tens of millions of cells with a similar number of vertices. In such a case, 
refining a mesh is not practical, since it would increase the computational effort 
and new vertices would not be placed on curved boundaries. A problem that arises 
for such large grids in the context how meshes have been treated in \texttt{deal.II} 
until now is that the coarse grid, i.e., potentially the whole mesh is shared by 
all processes. It might be a major difficulty in MPI-only parallelized applications 
on modern multi-core processors, since these applications might already run out of 
main memory during reading the mesh. Not even increasing the number of processes 
might help in this situation.

The new class \texttt{parallel::fullydistributed::Triangulation} targets this issue 
by distributing also the coarse grid, which is the reason for the name of the 
chosen namespace: it distributes the coarse grid as well as the refinement levels. Such 
a triangulation can be created by providing a \texttt{TriangulationDescription::Description} struct to each process, containing 
1) the relevant data to construct the local part of the coarse grid, 2) the 
translation of the local coarse-cell IDs to globally unique IDs, 3) the hierarchy 
of mesh refinements, and 4) the owner of the cells on the active mesh level as well 
as on the multigrid levels. Once the triangulation is set up with this struct, no 
changes to the mesh are allowed at the moment.

The \texttt{TriangulationDescription::Description} struct can be filled manually or 
by the utility functions from the \texttt{TriangulationDescription::Utilities} 
namespace. The function \texttt{create\_\allowbreak description\_\allowbreak 
from\_ \allowbreak triangulation()} can convert a base triangulation (partitioned 
serial \texttt{Tri\-angulation} and \texttt{parallel::distributed::Triangulation}) 
to such a struct. The advantage of this approach is that all known utility 
functions from the namespaces \texttt{GridIn} and \texttt{GridTools} can be used 
on the base triangulations before converting them to the structs. Since this 
function suffers from the same main memory problems as described above, we also 
provide the function \texttt{create\_description\_from\_triangulation\_in\_groups()}, 
which creates the structs only on the master process in a process group. These 
structs are filled one by one and are sent to the relevant processes once they are 
ready. A sensible process group size might contain all processes of one compute node.

The new (fully) distributed triangulation class works---in contrast to  
\texttt{parallel::distributed::\allowbreak Tri\-an\-gu\-la\-tion}---not only for 2D- and 3D- but also for 
1D-problems. It can be used in the context of geometric multigrid methods and 
supports periodic boundary conditions. Furthermore, hanging nodes are supported.

We intend to extend the usability of the new triangulation class in regard of 
different aspects, e.g., I/O. In addition, we would like to enable adaptive mesh 
refinement, a feature of the other triangulation classes, which is very much 
appreciated by many users. For repartitioning, we plan to use an oracle approach 
known from \texttt{parallel::distributed::Triangulation}. Here, we would like to 
rely on a user-provided partitioner, which might also be a graph partitioner.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Advances of the SIMD capabilities and the matrix-free infrastructure}
\label{subsec:mf}

%\begin{itemize}
%\item ECL
%\item VectorizedArrayType
%\end{itemize}

The class \texttt{VectorizedArray<Number>} is a key ingredient for the high 
node-level performance of the matrix-free algorithms in deal.II. It is a wrapper 
class around $n$ vector entries of type \texttt{Number} and delegates relevant 
function calls to appropriate Intrinsics instructions. Up to release 9.1, the 
vector length $n$ has been set at compile time of the library to the highest 
possible value supported by the given processor architecture.

The class \texttt{VectorizedArray} has been made more user-friendly by making 
it compatible with the STL algorithms found in the header \texttt{<algorithm>}. 
Now, it has following features:
\begin{itemize}
\item \texttt{VectorizedArray::size()} returns the vector length. This function 
replaces the public static attribute \texttt{VectorizedArray::n\_array\_elements}, 
which has been deprecated.
\item \texttt{VectorizedArray::value\_type} contains the underlying number type of 
the array.
\item \texttt{VectorizedArray} has an output operator 
\texttt{std::ostream\& operator<<(\&out, \&p)}.
\item \texttt{VectorizedArray::begin()} and \texttt{VectorizedArray::end()} allow 
range-based iteration over all vector entries.
\end{itemize}
Furthermore, the \texttt{VectorizedArray} class supports the following (tested) 
algorithms: \texttt{std::\allowbreak ad\-vance()}, \texttt{std::distance()}, and \texttt{std::max\_element()}.

It has been also extended with the second optional template argument 
\texttt{VectorizedArray<Number, size>} with \texttt{size} being related to the 
vector length, i.e., the number of lanes to be used and the instruction set to be 
used. By default, the number is set to the highest value supported by the given 
hardware, i.e., \texttt{VectorizedArray<double>} is translated on Skylake-based 
processors to \texttt{VectorizedArray<double, 8>}. A full list of supported 
vector lengths are presented in Table~\ref{tab:vectorizedarray}.

All matrix-free related classes (like \texttt{MatrixFree} and \texttt{FEEvaluation}) 
have been templated with the floating-point number type \texttt{Number} (e.g. \texttt{double} or \texttt{float}); the computations were performed implicitly 
on \texttt{VectorizedArray<Number>} structs with the highest 
instruction-set-architecture extension, with each lane responsible for a separate 
cell (vectoriziation over elements). In release 9.2, all matrix-free classes 
have been extended with a new optional template argument specifying the 
\texttt{VectorizedArrayType}. This allows users to select the vector length/ISA and, 
as a consequence, the number of cells to be processed at once. 
\begin{table}
\caption{Supported vector lengths of the class \texttt{VectorizedArray} and 
the corresponding instruction-set-architecture extensions. }\label{tab:vectorizedarray}
\centering
\begin{tabular}{ccc}
\toprule
\textbf{double} & \textbf{float} & \textbf{ISA}\\
\midrule
VectorizedArray<double, 1> & VectorizedArray<float, 1> & (auto-vectorization) \\
VectorizedArray<double, 2> & VectorizedArray<float, 4> & SSE2 \\ 
VectorizedArray<double, 4> & VectorizedArray<float, 8> & AVX/AVX2 \\ 
VectorizedArray<double, 8> & VectorizedArray<float, 16> & AVX-512 \\ 
\bottomrule
\end{tabular}

\caption{Comparison of relevant SIMD-related classes in deal.II and C++20.}\label{tab:simd}
\centering
\begin{tabular}{cc}
\toprule
\textbf{VectorizedArray (deal.II)} & \textbf{std::simd (C++20)} \\
\midrule
VectorizedArray<Number> & std::experimental::native\_simd<Number> \\
VectorizedArray<Number, size> & std::experimental::fixed\_size\_simd<Number, size> \\ \bottomrule
\end{tabular}
\end{table}

In standard (2D/3D) matrix-free applications with moderate polynomial degrees, 
we found that there is no reason to modify the default vector length of 
\texttt{VectorizedArray}, since it reaches the highest possible computational 
throughput despite of increased memory footprint. However, in the deal.II-based 
library \texttt{hyper.deal}~\cite{munch2020hyperdeal}, where the same matrix-free 
infrastructure was used for solving the 6D Vlasov--Poisson equation with high-order 
discontinuous Galerkin methods (with more than 1024 degrees of freedom per cell), the 
benefit of decreasing the number of cells processed by a single SIMD instruction was 
shown. That library works with \texttt{MatrixFree} objects of different SIMD-vector 
length in the same application, which would not have been possible before this 
release.

In a next step, we intend to support that users could set 
\texttt{VectorizedArrayType} to \texttt{Number}, which would lead to the usage not 
of \texttt{VectorizedArray<Number, 1>} but of a specialized code-path exploiting 
vectorization within an element~\cite{KronbichlerKormann2019}.

A side effect of introducing the new template argument \texttt{VectorizedArrayType} 
in the \texttt{MatrixFree} classes is that any data structures 
\texttt{VectorizedArrayType} can be processed if they support required 
functionalities like \texttt{size()} or \texttt{value\_type}. In this context, we 
would like to highlight that the new \texttt{C++20} feature \texttt{std::simd} 
can be processed by the matrix-free infrastructure with minor internal 
adjustment as an open pull request shows 
(see \url{https://github.com/dealii/dealii/pull/9994}).  
Table~\ref{tab:simd} gives a comparison of the deal.II-specific SIMD classes and 
the equivalent C++20 classes. We welcome the standardization of the SIMD 
parallelization paradigm in C++ and intend to replace step by step our own 
wrapper class, which has been continuously developed over the last decade. We 
would like to emphasize that the work invested in this class was not in vain, 
since many performance-relevant utility functions implemented with \texttt{VectorizedArray} in mind (e.g., \texttt{vectorized\_load\_and\_transpose} 
and \texttt{vectorized\_transpose\_and\_store}) will be still used, since they 
have not become part of the standard.

Further additions to the \texttt{MatrixFree} infrastructure consist of:
\begin{itemize}
\item a new variant of \texttt{MatrixFree::cell\_loop()}: It takes two
\texttt{std::function} objects with ranges on the locally owned degrees of freedom, one
with work to be scheduled before the cell operation first touches some
unknowns and another with work to be executed after the cell operation last 
touches them. The goal of
these functions is to bring vector operations close to the time when the
vector entries are accessed by the cell operation, which increases the cache
hit rate of modern processors by improved temporal locality. 
\item a new form of loop \texttt{MatrixFree::loop\_cell\_centric()}: This 
kind of loop can be used in the context of discontinuous Galerkin methods, 
where both cell and face integrals have to be evaluated. While in the case of 
the traditional \texttt{loop}, cell and face integrals have been performed 
independently, the new loop performs all cell and face integrals of a cell in 
one go. This includes that each face integral has to be evaluated twice, but 
entries have to be written into the solution vector only once with improved 
data locality. Previous publications based on \texttt{deal.II} have shown the 
relevance of the latter aspect for reaching higher performance. 
\end{itemize}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Advances in GPU support}
\label{subsec:gpu}

\begin{itemize}
\item overlapping of computation and communication in the case of CUDA-aware MPI
\end{itemize}

@@

In addition, there are seven new tutorial programs:
\begin{itemize}
\item \texttt{step-47}
\item \texttt{step-50}
\item \texttt{step-58}
\item \texttt{step-65} presents \texttt{TransfiniteInterpolationManifold}, a
manifold class that can propagate curved boundary information into the
interior of a computational domain, and \texttt{MappingQCache} for fast operations for
expensive manifolds.
\item \texttt{step-67} presents an explicit time integrator for the
compressible Euler equations discretized with a high-order discontinuous
Galerkin scheme using the matrix-free infrastructure. Besides the use of
matrix-free evaluators for systems of equations and over-integration, it also
presents \texttt{MatrixFreeOperators::CellwiseInverseMassMatrix}, a fast implementation
of the action of the inverse mass matrix in the DG setting using tensor
products. Furthermore, this tutorial demonstrates i.a. the usage of the new 
pre and post operations which can be passed to \texttt{cell\_loop()} 
(see also Subsection~\ref{subsec:mf}) and discusses performance-related aspects.
\item \texttt{step-69}
\item \texttt{step-70}
\end{itemize}

@@

The 9.2 release includes
\href{https://dealii.org/developer/doxygen/deal.II/changes_between_9_1_1_and_9_2_0.html}
     {around 15 incompatible changes}; see \cite{changes92}. The majority of these changes
should not be visible to typical user codes; some remove previously
deprecated classes and functions; and the majority change internal
interfaces that are not usually used in external
applications. However, some are worth mentioning:
\begin{itemize}
\item The functions:
\begin{itemize}
\item \texttt{DoFHandler::loccaly\_owned\_dofs\_per\_processor()}
\item \texttt{DoFHandler::loccaly\_owned\_mg\_dofs\_per\_processor()}
\end{itemize} have been deprecated. As discussed in Subsection~\ref{subsec:performance}, deal.II does not store information for all processes on all processes processes, but only the local information or the locally-relevant information. Users are asked to construct the global information on their own, e.g. by calling \texttt{Utilities::MPI::Allgather(locally\_owned\_info(), comm)}.
\item

% \item The \texttt{VectorView} class was removed. We recommend either copying the
%       vector subset into a \texttt{Vector} or using a \texttt{BlockVector}.
% \item The function \texttt{Subscriptor::subscribe()}, used through the
%   \texttt{SmartPointer} class, now requires a pointer to a
%       \texttt{std::atomic<bool>} that tracks whether or not the pointer to the
%       subscribed-to object is still valid.
% \item The \texttt{ConstraintMatrix} class gained a template parameter for the scalar
%       type and was been renamed \texttt{AffineConstraints}. Several methods that
%       take vectors or matrices as arguments,
%       such as \texttt{AffineConstraints::distribute\_local\_to\_global()},
%       now require that all matrix and vector arguments have matching number
%       types.
% \item Similarly, the functions \texttt{create\_mass\_matrix} and
%       \texttt{create\_boundary\_mass\_matrix} in the \texttt{MatrixCreator}
%       namespace no longer
%       support matrix and vector objects of different types.
\end{itemize}
